---
title: "Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Overview

**sciVizModules** extends [VizModules](https://cran.r-project.org/package=VizModules)
with a curated set of domain-specific Shiny modules for the biological, chemical,
and physical sciences. Each module wraps a plotting function with scientifically
meaningful defaults, axis conventions, and annotation layers, while preserving
the interactivity-first philosophy of the parent package.

Every module follows the same three-function contract:

- `*InputsUI(id, data, ...)` — the interactive controls.
- `*OutputUI(id)` — the (Plotly) plot output.
- `*Server(id, data, ...)` — the reactive logic that renders the plot.

Servers share a common signature, with `data` supplied as a **reactive**:

```r
someModuleServer(id, data, hide.inputs = NULL, hide.tabs = NULL, defaults = NULL)
```

Each module also ships a standalone `*App()` convenience function.

## The module gallery

To preview every module in one place, launch the bundled gallery app:

```{r gallery}
library(shiny)
runApp(system.file("apps/module-gallery", package = "sciVizModules"))
```

## Drop a module into your app

All modules follow the same pattern: `*InputsUI()` for controls, `*OutputUI()`
for the plot, and `*Server()` for the logic. Here is a minimal volcano plot
using the bundled `airway_deseq2` differential-expression results:

```{r volcano-example}
library(sciVizModules)

ui <- fluidPage(
    sidebarLayout(
        sidebarPanel(
            volcanoPlotInputsUI("de",
                airway_deseq2,
                defaults = list(
                    sig.thresh = 0.05,
                    fc.thresh = 1
                )
            )
        ),
        mainPanel(volcanoPlotOutputUI("de"))
    )
)

server <- function(input, output, session) {
    volcanoPlotServer("de", data = reactive(airway_deseq2))
}

shinyApp(ui, server)
```

## Try a module standalone

The fastest way to explore a module is its `*App()` function:

```{r apps}
if (interactive()) volcanoPlotApp()   # differential expression, airway_deseq2
if (interactive()) dittoDimPlotApp()  # single-cell reduction, example_sce
if (interactive()) doseResponseApp()  # pharmacology, dose_response
```

## Set defaults and hide controls

Pass a named `defaults` list to pre-fill inputs, and use `hide.inputs` /
`hide.tabs` to trim the control panel. Hidden inputs still pass their values to
the plot, so the module stays fully configured.

```{r defaults}
volcanoPlotServer(
    "de",
    data      = reactive(airway_deseq2),
    defaults  = list(sig.thresh = 0.01, fc.thresh = 1),
    hide.tabs = c("Legend/Scale")
)
```

Hidden inputs and tabs still feed their values into the plot, so the module
stays fully configured while exposing only what your users need.

## App factory with `createModuleApp()`

Every `*App()` convenience function (e.g. `volcanoPlotApp()`,
`survivalCurveApp()`, `dittoDimPlotApp()`) is a thin wrapper around VizModules'
`createModuleApp()`, which returns a full standalone app with data import, a
filterable data table, and dataset switching for any module:

```{r create-module-app}
library(sciVizModules)

app <- createModuleApp(
    inputs_ui_fn = volcanoPlotInputsUI,
    output_ui_fn = volcanoPlotOutputUI,
    server_fn    = volcanoPlotServer,
    data_list    = list("Airway DESeq2" = airway_deseq2),
    title        = "My Volcano Plot"
)
if (interactive()) runApp(app)
```

The data-frame modules take a `data_list`; the dittoSeq single-cell modules take
an `object_list` of `SingleCellExperiment` / `Seurat` / `SummarizedExperiment`
objects instead:

```{r dittoseq-app}
if (interactive()) {
    dittoDimPlotApp(object_list = list("Example SCE" = example_sce))
}
```

## Data format for each module

sciVizModules modules split into two families by input type: **data-frame**
modules and **dittoSeq / single-cell** modules. Every module below has a bundled
example dataset you can inspect (e.g. `head(airway_deseq2)`).

### Differential expression

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| Volcano plot | `volcanoPlotServer()` | `data.frame` | Effect-size column (e.g. `log2FoldChange`) and a significance column (e.g. `padj`) | `airway_deseq2` |
| MA plot | `maPlotServer()` | `data.frame` | Mean-abundance column (`baseMean` / `logCPM` / `AveExpr`) and an effect-size column (`log2FoldChange` / `logFC`) | `airway_deseq2` |

Column names differ between DE tools (`airway_deseq2`, `airway_edger`,
`airway_voom` are all bundled). The modules auto-detect common conventions, and
you can override the mapping via `defaults` (e.g.
`defaults = list(x.by = "logFC", y.by = "logCPM")`).

```{r de-format}
head(airway_deseq2)
#> baseMean, log2FoldChange, lfcSE, stat, pvalue, padj, symbol, ensembl
```

### GO / functional enrichment

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| Enrichment dot plot | `enrichmentDotPlotServer()` | `data.frame` | `clusterProfiler`-style table: a grouping column, a `Description` term column, a `"count/total"` `GeneRatio` string, and a p-value column | `example_enrichment` |
| GO fan (sunburst) plot | `goFanPlotServer()` | `data.frame` | GO enrichment results with a GO-ID column (auto-detected), an ontology (BP/CC/MF), and a numeric `fill` column (e.g. `qvalue`) | `example_enrichment` |

For the dot plot, dot size encodes the parsed `GeneRatio` and color encodes
\eqn{-\log_{10}(p)}.

```{r enrich-format}
head(example_enrichment)
#> Cluster, ONTOLOGY, ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, qvalue, Count
```

### Survival analysis

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| Kaplan-Meier curve | `survivalCurveServer()` | `data.frame` | Numeric follow-up `time` column and a `status` event column (`survival::Surv()` convention: 1 = censored, 2 = event); optional grouping column for stratified curves | `survival_lung` |

```{r surv-format}
head(survival_lung)
#> time, status, age, sex, ph.ecog
```

### Biochemical kinetics and pharmacology

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| Michaelis-Menten | `michaelisMentenServer()` | reactive `list` | A list with `data` (observed-points `data.frame`), `model` (fitted-line `data.frame`), and optional `stats` (an `nls` fit or named `K` / `Vmax` coefficients for annotation) | `mm_kinetics`, `mm_kinetics_line`, `mm_kinetics_fit` |
| Dose-response | `doseResponseServer()` | `data.frame` | A dose column (plotted on a log10 scale) and a response column; the module fits a log-logistic curve with **drc** | `dose_response` |

The Michaelis-Menten module consumes pre-computed pieces rather than a single
data frame:

```{r mm-format}
str(list(
    data  = mm_kinetics,       # columns S (substrate), v (velocity)
    model = mm_kinetics_line,  # predicted S, v over a fine grid
    stats = mm_kinetics_fit    # nls fit -> supplies K and Vmax
))
```

```{r dr-format}
head(dose_response)
#> dose_uM, rep1_response_pct, ..., mean_response_pct, sd_response_pct
```

### Single-cell / RNA-seq (dittoSeq wrappers)

These modules wrap `dittoSeq` plotting functions. They all accept a
**`SingleCellExperiment`, `Seurat`, or `SummarizedExperiment`** object and take
`object_list` (rather than `data_list`) in their `*App()` functions. The bundled
`example_sce` (a simulated `SingleCellExperiment` with `PCA`/`TSNE`/`UMAP`
reductions and both discrete and continuous metadata) works with all of them.

All of these take the same **data type**: a `SingleCellExperiment`, `Seurat`,
or `SummarizedExperiment` object.

| Module | Server function | Purpose |
|---|---|---|
| Dimensional reduction | `dittoDimPlotServer()` | Cells in a reduction (UMAP/t-SNE/PCA), colored by a gene or metadata |
| Hex-binned reduction | `dittoDimHexServer()` | Hex-binned version of the dim plot for dense embeddings |
| Scatter | `dittoScatterPlotServer()` | Two continuous variables, optional third for color |
| Distribution | `dittoPlotServer()` | A continuous variable across groups (jitter / violin / box / ridge) |
| Ridge + jitter | `dittoRidgeJitterServer()` | Ridge plots with overlaid jitter |
| Composition bar | `dittoBarPlotServer()` | Discrete composition per group (percent or count) |
| Frequency | `dittoFreqPlotServer()` | Per-sample frequencies of a discrete variable across groups |

```{r sce-format}
data(example_sce)
example_sce
```

## Which plot parameters are exposed?

Modules wrap plotting functions from **dittoViz**, **plotthis**, **dittoSeq**,
and several specialized packages. To see which arguments a module surfaces:

1. Open the module input help page, e.g., `?volcanoPlotInputsUI` or
   `?dittoDimPlotInputsUI`. The **Details** section notes which arguments from
   the underlying plot function are wired through and any that are intentionally
   omitted.
2. Cross-reference the base plot documentation (`?dittoSeq::dittoDimPlot`, etc.).
   Input names in `defaults` line up with those function arguments when they are
   supported.

If an argument is listed as missing or non-functional in the module docs, it has
been intentionally hidden because it does not round-trip well in the interactive
Plotly output.
