---
title: "Quick Start with sciVizModules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start with sciVizModules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Overview

**sciVizModules** extends [VizModules](https://cran.r-project.org/package=VizModules)
with a curated set of domain-specific Shiny modules for the biological, chemical,
and physical sciences. Each module wraps a plotting function with scientifically
meaningful defaults, axis conventions, and annotation layers, while preserving
the interactivity-first philosophy of the parent package.

Every module follows the same three-function contract:

- `*InputsUI(id, data, ...)` — the interactive controls.
- `*OutputUI(id)` — the (Plotly) plot output.
- `*Server(id, data, ...)` — the reactive logic that renders the plot.

Servers share a common signature, with `data` supplied as a **reactive**:

```r
someModuleServer(id, data, hide.inputs = NULL, hide.tabs = NULL, defaults = NULL)
```

Each module also ships a standalone `*App()` convenience function.

## The module gallery

To preview every module in one place, launch the bundled gallery app:

```{r gallery}
library(shiny)
runApp(system.file("apps/module-gallery", package = "sciVizModules"))
```

## Drop a module into your app

Here is a minimal volcano-plot app driven by the bundled `airway_deseq2` results:

```{r volcano-example}
library(sciVizModules)

ui <- fluidPage(
    sidebarLayout(
        sidebarPanel(
            volcanoPlotInputsUI("de", airway_deseq2)
        ),
        mainPanel(volcanoPlotOutputUI("de"))
    )
)

server <- function(input, output, session) {
    volcanoPlotServer("de", data = reactive(airway_deseq2))
}

shinyApp(ui, server)
```

## Try a module standalone

The fastest way to explore a module is its `*App()` function:

```{r apps}
if (interactive()) volcanoPlotApp()   # differential expression, airway_deseq2
if (interactive()) dittoDimPlotApp()  # single-cell reduction, example_sce
if (interactive()) doseResponseApp()  # pharmacology, dose_response
```

## Set defaults and hide controls

Pass a named `defaults` list to pre-fill inputs, and use `hide.inputs` /
`hide.tabs` to trim the control panel. Hidden inputs still pass their values to
the plot, so the module stays fully configured.

```{r defaults}
volcanoPlotServer(
    "de",
    data      = reactive(airway_deseq2),
    defaults  = list(sig.thresh = 0.01, fc.thresh = 1),
    hide.tabs = c("Legend/Scale")
)
```

## Module data formats

| Module family | Modules | Expected data |
|---|---|---|
| Differential expression | `volcanoPlot`, `maPlot` | `data.frame` with effect-size and significance columns |
| Enrichment | `enrichmentDotPlot`, `goFanPlot` | `clusterProfiler`-style enrichment `data.frame` |
| Survival | `survivalCurve` | `data.frame` with `time` and `status` columns |
| Pharmacology / kinetics | `doseResponse`, `michaelisMenten` | dose/response `data.frame`; kinetics list |
| Copy number | `cnSegmentPlot` | a `CNSegment` object from `sesame::cnSegmentation()` |
| Single cell (dittoSeq) | `dittoDimPlot`, `dittoScatterPlot`, `dittoPlot`, `dittoBarPlot`, `dittoDimHex`, `dittoFreqPlot`, `dittoRidgeJitter` | `SingleCellExperiment` / `Seurat` / `SummarizedExperiment` |

Each module ships a bundled example dataset (e.g. `airway_deseq2`,
`example_enrichment`, `survival_lung`, `dose_response`, `example_sce`) so you can
try it immediately.

## Session info

```{r session-info, eval = TRUE}
sessionInfo()
```
