---
title: "Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Every chunk below either launches a Shiny app or needs a Bioconductor
# dependency that is only Suggested, so none of them run at build time.
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

## Overview

**sciVizModules** extends [VizModules](https://github.com/j-andrews7/VizModules)
with a curated set of domain-specific Shiny modules for the biological, chemical,
and physical sciences. Each module wraps a plotting function with scientifically
meaningful defaults, axis conventions, and annotation layers, while preserving
the interactivity-first philosophy of the parent package.

Every module follows the same three-function contract:

- `*InputsUI(id, data, ...)` — the interactive controls.
- `*OutputUI(id)` — the (Plotly) plot output.
- `*Server(id, data, ...)` — the reactive logic that renders the plot.

Servers share a common signature, with `data` supplied as a **reactive**:

```r
someModuleServer(id, data, hide.inputs = NULL, hide.tabs = NULL, defaults = NULL)
```

Each module also ships a standalone `*App()` convenience function.

Note that this package sets no `LazyData`, so the bundled example datasets need
a `data()` call before you can use them by name:

```{r load-data}
library(sciVizModules)
data(airway_deseq2)
data(example_sce)
data(survival_lung)
```

## Working with an AI coding agent

sciVizModules builds on VizModules, which ships three agent skills describing
its APIs. Install them into your project once and any compatible agent (Claude
Code, GitHub Copilot, OpenAI Codex) will discover them:

```{r skills}
VizModules::use_vizmodules_skills(".", client = "claude")  # or "agents" / "copilot"
```

Point your agent at the one that matches the job:

- **Putting sciVizModules modules into an app** (`defaults`, `hide.inputs` /
  `hide.tabs`, `createModuleApp()`, the data filter, the Figure Builder,
  source-data export): `vizmodules-app`. It also carries a generated inventory
  of every VizModules module's column-mapping keys, colour key and tab names,
  which the wrapper modules here inherit.
- **Building your own module on top of one of these, or of a VizModules base
  module** (adding inputs, transforming the data before it is plotted, driving
  the wrapped module's parameters): `vizmodules-custom-module`. Most of
  sciVizModules is built exactly this way, so its modules are good worked
  examples.
- **Writing a module from scratch around your own plotting function**:
  `vizmodules-custom-module` for the Shiny wiring, plus `vizmodules-new-module`
  for the full `*InputsUI()` / `*OutputUI()` / `*Server()` / `*App()` layout,
  documentation and tests. The latter is written for contributions to
  VizModules itself, but the structure carries over.

The skills are copied from your installed VizModules, so re-run the call with
`overwrite = TRUE` after upgrading it. `AGENTS.md` in the sciVizModules
repository covers the conventions specific to this package on top of those.

## The module gallery

To preview every module in one place, launch the bundled gallery app:

```{r gallery}
library(shiny)
runApp(system.file("apps/module-gallery", package = "sciVizModules"))
```

## Drop a module into your app

All modules follow the same pattern: `*InputsUI()` for controls, `*OutputUI()`
for the plot, and `*Server()` for the logic. Here is a minimal volcano plot
using the bundled `airway_deseq2` differential-expression results:

```{r volcano-example}
library(sciVizModules)

ui <- fluidPage(
    sidebarLayout(
        sidebarPanel(
            volcanoPlotInputsUI("de",
                airway_deseq2,
                defaults = list(
                    sig.thresh = 0.05,
                    fc.thresh = 1
                )
            )
        ),
        mainPanel(volcanoPlotOutputUI("de"))
    )
)

server <- function(input, output, session) {
    volcanoPlotServer("de", data = reactive(airway_deseq2))
}

shinyApp(ui, server)
```

## Try a module standalone

The fastest way to explore a module is its `*App()` function:

```{r apps}
if (interactive()) volcanoPlotApp()     # differential expression, airway_deseq2
if (interactive()) dittoDimPlotApp()    # single-cell reduction, example_sce
if (interactive()) doseResponseApp()    # pharmacology, dose_response
if (interactive()) cnSegmentPlotApp()   # copy number, example_cn_segment
if (interactive()) forestPlotApp()      # Cox / logistic / linear model, survival_lung
if (interactive()) pcaBiplotApp()       # PCAtools pca object, example_pca
if (interactive()) alphafoldConfidenceApp() # AlphaFold DB prediction for p53
if (interactive()) manhattanPlotApp()   # GWAS summary statistics, example_gwas
if (interactive()) gseaEnrichmentPlotApp() # fgsea example ranks and pathways
if (interactive()) crisprScreenRankApp() # simulated MAGeCK gene summary
```

## Set defaults and hide controls

Pass a named `defaults` list to pre-fill inputs, and use `hide.inputs` /
`hide.tabs` to trim the control panel. Hidden inputs still pass their values to
the plot, so the module stays fully configured.

```{r defaults}
volcanoPlotServer(
    "de",
    data        = reactive(airway_deseq2),
    defaults    = list(sig.thresh = 0.01, fc.thresh = 1),
    hide.inputs = c("shape.by"),
    hide.tabs   = c("Legend")
)
```

Hidden inputs and tabs still feed their values into the plot, so the module
stays fully configured while exposing only what your users need.

The tab names come from the module being wrapped, so check its help page before
naming one: a `hide.tabs` entry that matches nothing is silently ignored. For
the volcano and MA modules the tabs are those of
`VizModules::dittoViz_scatterPlotInputsUI()` -- Data, Adjustments, Points,
Colors, Facet, Annotations, Legend, Trajectory, Lines, Axes, Plotly, Extras.
The `vizmodules-app` agent skill (see above) lists the tab names and `defaults`
keys of every VizModules base module in its module inventory.

Group colours are a `defaults` entry too. The volcano and MA modules assemble
`color.up` / `color.down` / `color.ns` into the wrapped module's `color.panel`
mapping, so the colours open where you asked and stay editable on its Colors
tab:

```{r de-colors}
volcanoPlotServer(
    "de",
    data     = reactive(airway_deseq2),
    defaults = list(color.up = "#D55E00", color.down = "#0072B2")
)
```

## App factory with `createModuleApp()`

Every `*App()` convenience function (e.g. `volcanoPlotApp()`,
`survivalCurveApp()`, `dittoDimPlotApp()`) is a thin wrapper around VizModules'
`createModuleApp()`, which returns a full standalone app with data import, a
filterable data table, and dataset switching for any module:

```{r create-module-app}
library(sciVizModules)

app <- VizModules::createModuleApp(
    inputs_ui_fn = volcanoPlotInputsUI,
    output_ui_fn = volcanoPlotOutputUI,
    server_fn    = volcanoPlotServer,
    data_list    = list("Airway DESeq2" = airway_deseq2),
    title        = "My Volcano Plot"
)
if (interactive()) runApp(app)
```

The `vizmodules-app` agent skill covers `createModuleApp()` and its data filter
in more depth.

The data-frame modules take a `data_list`; the dittoSeq single-cell modules take
an `object_list` of `SingleCellExperiment` / `Seurat` / `SummarizedExperiment`
objects instead:

```{r dittoseq-app}
if (interactive()) {
    dittoDimPlotApp(object_list = list("Example SCE" = example_sce))
}
```

## Data format for each module

sciVizModules modules split into two families by input type: **data-frame**
modules and **dittoSeq / single-cell** modules. Every module below has a bundled
example dataset you can inspect (e.g. `head(airway_deseq2)`).

### Differential expression

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| Volcano plot | `volcanoPlotServer()` | `data.frame` | Effect-size column (e.g. `log2FoldChange`) and a significance column (e.g. `padj`) | `airway_deseq2` |
| MA plot | `maPlotServer()` | `data.frame` | Mean-abundance column (`baseMean` / `logCPM` / `AveExpr`) and an effect-size column (`log2FoldChange` / `logFC`) | `airway_deseq2` |

Column names differ between DE tools (`airway_deseq2`, `airway_edger`,
`airway_voom` are all bundled). The modules auto-detect common conventions, and
you can override the mapping via `defaults` (e.g.
`defaults = list(x.by = "logFC", y.by = "logCPM")`).

```{r de-format}
head(airway_deseq2)
#> baseMean, log2FoldChange, lfcSE, stat, pvalue, padj, symbol, ensembl
```

### GO / functional enrichment

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| Enrichment dot plot | `enrichmentDotPlotServer()` | `data.frame` | `clusterProfiler`-style table: a grouping column, a `Description` term column, a `"count/total"` `GeneRatio` string, and a p-value column | `example_enrichment` |
| GO fan (sunburst) plot | `goFanPlotServer()` | `data.frame` | GO enrichment results with a GO-ID column (auto-detected), an ontology (BP/CC/MF), and a numeric `fill` column (e.g. `qvalue`) | `example_enrichment` |

For the dot plot, dot size encodes the parsed `GeneRatio` and color encodes
\eqn{-\log_{10}(p)}.

```{r enrich-format}
head(example_enrichment)
#> Cluster, ONTOLOGY, ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, qvalue, Count
```

### Survival analysis

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| Kaplan-Meier curve | `survivalCurveServer()` | `data.frame` | Numeric follow-up `time` column and a `status` event column (`survival::Surv()` convention: 1 = censored, 2 = event); optional grouping column for stratified curves | `survival_lung` |

```{r surv-format}
head(survival_lung)
#> time, status, age, sex, ph.ecog
```

### Clinical models

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| Forest plot | `forestPlotServer()` | `data.frame` | Raw data: for a Cox model a time and an event column (as for the survival curve); for a logistic model a binary outcome; for a linear model a numeric outcome; plus any numeric or categorical covariates. The module fits the model | `survival_lung` |

```{r forest-format}
forestPlot(survival_lung,
    model = "cox", time = "time", status = "status",
    covariates = c("age", "sex", "ph.ecog")
)
```

### Biochemical kinetics and pharmacology

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| Michaelis-Menten | `michaelisMentenServer()` | reactive `list` | A list with `data` (observed-points `data.frame`), `model` (fitted-line `data.frame`), and optional `stats` (an `nls` fit or named `K` / `Vmax` coefficients for annotation) | `mm_kinetics`, `mm_kinetics_line`, `mm_kinetics_fit` |
| Dose-response | `doseResponseServer()` | `data.frame` | A dose column (plotted on a log10 scale) and a response column; the module fits a log-logistic curve with **drc** | `dose_response` |

The Michaelis-Menten module consumes pre-computed pieces rather than a single
data frame:

```{r mm-format}
str(list(
    data  = mm_kinetics,       # columns S (substrate), v (velocity)
    model = mm_kinetics_line,  # predicted S, v over a fine grid
    stats = mm_kinetics_fit    # nls fit -> supplies K and Vmax
))
```

```{r dr-format}
head(dose_response)
#> dose_uM, rep1_response_pct, ..., mean_response_pct, sd_response_pct
```

### Copy number

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| Copy-number segments | `cnSegmentPlotServer()` | `CNSegment`, or a named list of them | A `CNSegment` object as returned by `sesame::cnSegmentation()`: bin-level log2 signal ratios in `bin.signals`, their coordinates in `bin.coords`, and a `genomeInfo` carrying `cytoBand` (for chromosome ticks and centromere guides) and optionally `genes` (for labels) | `example_cn_segment` |

Passing a named list plots one panel per sample, stacked over a shared genomic
x-axis. Like the PCAtools and AlphaFold modules, it does not take a data frame,
so it cannot go in the Figure Builder.

```{r cn-format}
data(example_cn_segment)
example_cn_segment
```

### Genetics (GWAS)

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| Manhattan plot | `manhattanPlotServer()` | `data.frame` | Summary statistics with chromosome, position and p-value columns (and optionally a variant id), named in any common style (`CHR`/`BP`/`P`, `chromosome`/`base_pair_location`/`p_value`, ...) | `example_gwas` |
| GWAS QQ plot | `gwasQQPlotServer()` | `data.frame` | Summary statistics with a p-value column | `example_gwas` |

### Gene set enrichment (GSEA)

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| GSEA enrichment plot | `gseaEnrichmentPlotServer()` | fgsea bundle or `gseaResult` | `list(stats = <named numeric>, pathways = <named list>, results = <fgsea table, optional>)`, or a clusterProfiler `gseaResult` | `example_gsea` |

### Functional genomics screens

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| CRISPR screen rank | `crisprScreenRankServer()` | `data.frame` | A MAGeCK RRA `gene_summary.txt`, read with `read_mageck()` | simulated, in `inst/extdata` |

```{r crispr-format}
screen <- read_mageck(system.file("extdata", "example_mageck.gene_summary.txt.gz",
    package = "sciVizModules"))
head(screen)
```

### Principal component analysis (PCAtools)

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| Biplot | `pcaBiplotServer()` | PCAtools `pca` | The object `PCAtools::pca()` returns: sample scores (`rotated`), variable `loadings`, `variance` explained, and sample `metadata` | `example_pca` |
| Scree plot | `pcaScreePlotServer()` | PCAtools `pca` | As above | `example_pca` |
| Loadings plot | `pcaLoadingsPlotServer()` | PCAtools `pca` | As above | `example_pca` |
| Pairs plot | `pcaPairsPlotServer()` | PCAtools `pca` | As above | `example_pca` |
| PC-metadata correlation | `pcaEigencorPlotServer()` | PCAtools `pca` | As above; needs sample `metadata` | `example_pca` |

Their `*App()` functions take a named `pca_list`. A `pca` object is a plain
list, so the modules read it without PCAtools installed; PCAtools is needed
only to make one, and to mark the elbow on the scree plot.

```{r pca-format}
data(example_pca)
round(example_pca$variance, 1)
head(example_pca$metadata)
```

### Structural biology

| Module | Server function | Data type | Expected data | Example data |
|---|---|---|---|---|
| AlphaFold confidence | `alphafoldConfidenceServer()` | `read_alphafold()` result | AlphaFold DB `*-predicted_aligned_error_v*.json` and `*-confidence_v*.json`, a ColabFold `*_scores_*.json`, or a predicted model (`.pdb` / `.cif`) whose B-factors hold the pLDDT; files may be gzipped | AlphaFold DB prediction for p53 (P04637), in `inst/extdata` |

```{r af-format}
af <- read_alphafold(
    pae = system.file("extdata", "AF-P04637-F1-predicted_aligned_error_v6.json.gz",
        package = "sciVizModules"),
    confidence = system.file("extdata", "AF-P04637-F1-confidence_v6.json.gz",
        package = "sciVizModules")
)
af
alphafoldConfidence(af)
```

AlphaFold DB data is available under CC-BY 4.0; cite Jumper et al. (2021) and
Varadi et al. (2024) when you use it.

### Single-cell / RNA-seq (dittoSeq wrappers)

These modules wrap `dittoSeq` plotting functions. They all accept a
**`SingleCellExperiment`, `Seurat`, or `SummarizedExperiment`** object and take
`object_list` (rather than `data_list`) in their `*App()` functions. The bundled
`example_sce` (a simulated `SingleCellExperiment` with `PCA`/`TSNE`/`UMAP`
reductions and both discrete and continuous metadata) works with all of them.

All of these take the same **data type**: a `SingleCellExperiment`, `Seurat`,
or `SummarizedExperiment` object.

| Module | Server function | Purpose |
|---|---|---|
| Dimensional reduction | `dittoDimPlotServer()` | Cells in a reduction (UMAP/t-SNE/PCA), colored by a gene or metadata |
| Hex-binned reduction | `dittoDimHexServer()` | Hex-binned version of the dim plot for dense embeddings |
| Scatter | `dittoScatterPlotServer()` | Two continuous variables, optional third for color |
| Distribution | `dittoPlotServer()` | A continuous variable across groups (jitter / violin / box / ridge) |
| Ridge + jitter | `dittoRidgeJitterServer()` | Ridge plots with overlaid jitter |
| Composition bar | `dittoBarPlotServer()` | Discrete composition per group (percent or count) |
| Frequency | `dittoFreqPlotServer()` | Per-sample frequencies of a discrete variable across groups |

```{r sce-format}
data(example_sce)
example_sce
```

## Multi-panel figures

The data-frame modules are registered with the VizModules Figure Builder, so
several panels can be dragged, sized and exported as one figure:

```{r figure-builder}
sciFigureBuilderApp()
```

`sci_figure_builder_registry()` returns the registry itself, in the shape
`VizModules::figureBuilderServer()` documents, so you can extend it with your
own modules:

```{r figure-builder-registry}
registry <- sci_figure_builder_registry()
names(registry)
str(registry$volcano, max.level = 1)
```

The `vizmodules-app` agent skill covers embedding the Figure Builder in your own
app, and `vizmodules-custom-module` covers building a module to register in it.

The dittoSeq modules, `cnSegmentPlot`, `michaelisMenten`, the PCAtools modules,
`gseaEnrichmentPlot` and `alphafoldConfidence` are not registered: the
builder's dataset catalogue holds data frames, and those take other objects.
`forestPlot`, the GWAS modules and `crisprScreenRank` are not registered yet.

## Export the source data

Every module's control panel ends in a **Source Data** button. It downloads a
ZIP holding a self-contained interactive HTML of the plot, an SVG and a PNG of
it as drawn on screen, a CSV of the data actually plotted, a CSV of the UI input
values, and a statistics table where the module computes one. The Figure Builder
writes one such set per panel, named after the panel label.

This is plumbed by `VizModules::collect_source_data()` and
`VizModules::create_source_download_handler()`, which every module here already
wires up, so nothing is needed on your side. Two notes:

- The SVG and PNG are photographed in the browser off the graph you are looking
  at, so they carry anything applied after the figure was built. The button
  therefore pauses briefly before the archive arrives.
- The interactive HTML needs pandoc (via `htmlwidgets::saveWidget()`). Without
  it the archive still downloads, with a warning and no HTML.

## Which plot parameters are exposed?

Modules wrap plotting functions from **dittoViz**, **plotthis**, **dittoSeq**,
and several specialized packages. To see which arguments a module surfaces:

1. Open the module input help page, e.g., `?volcanoPlotInputsUI` or
   `?dittoDimPlotInputsUI`. The **Details** section notes which arguments from
   the underlying plot function are wired through and any that are intentionally
   omitted.
2. Cross-reference the base plot documentation (`?dittoSeq::dittoDimPlot`, etc.).
   Input names in `defaults` line up with those function arguments when they are
   supported.

If an argument is listed as missing or non-functional in the module docs, it has
been intentionally hidden because it does not round-trip well in the interactive
Plotly output.
